---
title: "Data Science Task"
author: "Ishwinder Kaur"
date: May 1, 2020
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: paper
    toc_float: true
    number_sections: true
    self_contained: yes
    highlight: pygments
    code_folding: hide
---

# Introduction

This R markdown demonstrates predictive analysis for time series data.
For this analysis we use the <b> Bowel Cancer Screening dataset.</b>

```{r warning=F, message=F, error=F}

#loading the relevant libraries
library(tidyverse)
#library(ts)
library(ggplot2)
library(forecast)

options(scipen = 999)

kLightBlue = "#00AAE5"
kDarkBlue = "#005AAD"
kPurple = "#A33299"


```

# About the dataset 

This dataset is taken from the data provided as part of the <b>National Bowel Cancer Screening Program</b>. The data consists of number of screening kits returned, by sex and quarter, people aged 50–74, 1 Jan–31 Mar 2014 to 1 Apr–30 Jun 2019.

Link: https://www.aihw.gov.au/reports/cancer-screening/national-cancer-screening-programs-participation/data

For the purpose of this analysis, we will use the total number of persons.

```{r}

# about the dataset:

bowel = readxl::read_xlsx('bowel screening data.xlsx')

# converting bowel screen dataset into a timeseries
bowel_ts <- ts(bowel[,'Persons'], start=2014, frequency=4)

# plotting the time series
autoplot(bowel_ts) +
  ggtitle("Bowel Screening Kits returned") +
  xlab("Year") +
  ylab("Number of screens") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme_bw()

```

# Main model and forecasts

The first model is a linear model. 

Below is the summary of the model:

```{r}
fit.cancer <- tslm(bowel_ts ~ trend + season)
summary(fit.cancer)

```


## Plot of the actual and fitted values

Below is a plot of the actual versus fitted values.

```{r}
cbind(Time = time(bowel_ts), Actuals=bowel_ts, Fitted=fitted(fit.cancer)) %>%
  as.data.frame() %>%
  gather(key = 'Type', value = 'value', -Time) %>% 
  ggplot(aes(x = Time, 
             y = value,
             colour = as.factor(Type) )) +
  ggtitle("Bowel Screening Kits returned/forecasts") +
  geom_point() +
  geom_line() +
  ylab("Actual/Fitted") + xlab("Time") +
  ggtitle("Quarterly bowel screen kits returned") +
  theme_bw() +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.2, 0.85)
  ) +
  scale_y_continuous(limits = c(0,500000), expand = c(0,0), labels = scales::comma)
#  scale_colour_brewer(palette="Dark2", name="Quarter") +
#  geom_abline(intercept=0, slope=1) 
  
```

## Forecasts 

Following are the forecasts for the model created above for the 10 quarters starting from 2019 Q3 to 2021 Q4.

```{r}
fcast <- forecast(fit.cancer)
autoplot(fcast) +
  ggtitle("Forecasts of bowel screens") +
  xlab("Time") + ylab("Number of screens") +
  ggtitle("Quarterly bowel screen kits returned with forecasts") +
  theme_bw() +
  theme_classic() +
  scale_y_continuous(limits = c(0,500000), expand = c(0,0), labels = scales::comma)


```

## Accuracy

Accuracy of the main model is as follows:

```{r}
accuracy(fcast)
```


# Other models

##  Alternative model 1

Alternative Model 1: Exponential smoothing

```{r}

fc1 = ses(bowel_ts, h = 4)
round(accuracy(fc1),2)
```

##  Alternative model 2

Alternative Model 2: Forecasting with Holt’s Method

```{r}
fc2 = holt(bowel_ts, h = 2)
round(accuracy(fc2),2)
```

In comparison to the alternative models, the first model still has higher accuracy and lower RMSE.